Modeling Lake Trophic State: A Data Mining Approach

### 2014-09-12 inst/doc/manuscript.Rmd
monkeying with latex

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author:
- affiliation: aed
  email: hollister.jeff(at)epa.gov
  footnote: Corresponding author
  name: Jeffrey W. Hollister
- affiliation: aed
  name: W. Bryan Milstead
- affiliation: aed
  name: Betty J. Kreakie
address:
- address: "US Environmental Protection Agency Office of Research and Development National
    Health and Environmental Effects Research Laboratory Atlantic Ecology Division 27
    Tarzwell Drive  Narragansett, RI, 02879, USA  "
  code: aed
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/elsarticle.latex
  html_document: null
  word_document: null
fontsize: 11pt
csl: components/ecology.csl
documentclass: elsarticle
layout: review
linenumbers: no
bibliography: components/LakeTrophicModelling.bib
abstract:  Productivity of lentic ecosystems has been well studied and it is 
  widely accepted that as nutrient inputs increase,  productivity increases and
  lakes transition from low trophic state (e.g. oligotrophic) to higher trophic 
  states  (e.g. eutrophic). These broad trophic state classifications are good 
  predictors of ecosystem health and ecosystem services/disservices (e.g. recreation,
  aesthetics, fisheries, and harmful algal blooms). While the relationship between 
  nutrients and trophic state provides reliable predictions, it requires *in situ* 
  water quality data in order to paramterize the model.  This limits the application 
  of these models to lakes with existing and, more importanly, available water quality
  data.  To expand our ability to predict in lakes without water quality data, we take
  advantage of the availability of a large national lakes water quality database, 
  land use/land cover data, lake morphometry data, other universally available data, 
  and modern data mining approaches to build and assess models of lake tropic state 
  that may be more universally applied.  We use random forests and random forest 
  variable selection to identify variables to be used for predicting trophic state and 
  we compare the performance of two models of trophic state (as determined by 
  chlorophyll a concentration).   The first model estimates trophic state with *in situ* 
  as well as universally available data and the second model uses universally available 
  data only. For each of these models we used three separate trophic state categories, 
  for a total of six models. Overall accuracy for the *in situ* and universal data 
  models ranged from xx% to xx% and xx, xx, and xx described the most variation in 
  trophic state.  For the universal data only models, Overall accuraccy ranged from xx% 
  to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it 
  is believed that the presence and abundance of cyanobacteria is strongly associated 
  with trophic state.  To test this we examine the association between estimates of 
  cyanobacteria biovolume and the measured and predicted trophic state.  Expanding 
  these preliminary results to include cyanobacteria taxa indicates that cyanobacteria 
  are significantly more likely to be found in highly eutrophic lakes. These results 
  suggest that predictive models of lake trophic state may be improved with additional 
  information on the landscape surrounding lakes and that those models provide 
  additional information on the presence of potentially harmful cyanobacteria taxa. 
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

@@

knit_hooks$set(tab.cap = function(before, options, envir) {
                  if(!before) { 
                    paste('\n\n:', options$tab.cap, sep='') 
                  }
                })
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == FALSE) {
    x
  } else
    default_output_hook(x,options)
})

```

#Introduction

Productivity in lentic systems is often categorized across a range of tropic states (e.g. the tropic continuum) from early succesional (i.e. oligotrophic)to late successional lakes (i.e. hypereutrophic) [@carlson1977trophic].  Lakes naturally occur across the range of trophic state and higher primary productivity is not necessarily a predictor of poor ecological condition.  Lakes that are naturally oligotrophic occur in nutrient poor areas or have a more recent geologic history.  These lakes are often found in higher elevations, have clear water, and are often favored for drinking water or direct contact recreation (e.g. swimming).   Lakes with higher productivity (e.g. eutrophic lakes) have greater nutrient loads, tend to be less clear, have greater density of aquatic plants, and often support more diverse and abundant fish communities.  Lakes will naturally shift to higher trophic states but this is a slow process.  Given this fact, monitoring trophic state allows the identification of rapid shifts in trophic state or locating lakes with unusually high productivity (e.g. hypereutrophic).  These cases are indicative of lakes under greater anthropogenic nutrient loads, also known as cultural eutrophication, and are more likely to be at risk of fish kills, fouling, and harmful algal blooms[@smith1998cultural;@smith1999eutrophication;@smith2006eutrophication].  Given the association between trophic state and many ecosystem services and disservices, being able to model trophic state could allow for estimating trophic state in unmonitored lakes and provide a first cut at identifying lakes with the potential for harmful algal blooms and other problems associated with cultural eutrophication.

@@

##Data and Study Area

We utilize four primary sources of data for this study,the National Lakes Assessment (NLA), the National Lake Cover Dataset (NLCD), modeled lake morphometery,  and estimated cyanobacteria biovolumes [@usepa2009national;@homer2004development;@xian2009updating;@hollister2010volume;@hollister_predicting_2011;@lakemorpho2014;@beaulieu2013nutrients].  All datasets are national in scale and provide a unique snapshot view of the condition of lakes in the United States'.

@@

##Predicting Trophic State with Random Forests

Random forest is a machine learning algorithm that aggregates numerous decision trees in order to obtain a consensus prediction of the response categories [@breiman2001random].  Bootstrapped sample data is recursively partitioned according to a given random subset of predictor variables and completely grown without pruning.   With each new tree, both the sample data and predictor variable subset is randomly selected.

@@

##Model Details

Using a combination of the `varSelRF` and `randomForest` we ran models for six combinations of variables and trophic state classifications.  These combinations included different combinations of the Chlorphyll *a* trophic states (Table \ref{tab:trophicStateTable}) along with all variables and the GIS only variables (i.e. no *in situ* infromation).  The six model combinations were:

@@

#Results

```{r figSetup,echo=FALSE,eval=TRUE,cache=TRUE}
#create list of all selected variables and assign a color
vars<-ls(pattern="_vars")
all_vars<-vector()
for(i in vars){
  all_vars<-c(all_vars,get(i))
}
all_vars<-unique(all_vars)

@@

##Model 1: 4 Trophic States ~ All Variables

```{r ALL_TS4, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts4<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_4,numCore=4,100,
    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="all_ts4",time=TRUE)
```

>>>>

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/elsarticle.latex
  html_document: null
  word_document: null
fontsize: 11pt
capsize: normalsize
csl: components/ecology.csl
documentclass: elsarticle
spacing: doublespacing
linenumbers: no
bibliography: components/LakeTrophicModelling.bib
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

@@

knit_hooks$set(tab.cap = function(before, options, envir) {
                  if(!before) { 
                    paste('\n\n:', options$tab.cap, sep='') 
                  }
                })
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == FALSE) {
    x
  } else
    default_output_hook(x,options)
})
```
\begin{center} 
Jeffrey W. Hollister*, W. Bryan Milstead, Betty J. Kreakie

*US Environmental Protection Agency, Office of Research and Development, National Health and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell Drive  Narragansett, RI, 02879, USA*

------
*corresponding author:hollister.jeff@epa.gov
\end{center}

*#Abstract
Productivity of lentic ecosystems has been well studied and it is widely accepted that as nutrient inputs increase,  productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states  (e.g. eutrophic). These broad trophic state classifications are good predictors of ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms). While the relationship between  nutrients and trophic state provides reliable predictions, it requires *in situ* water quality data in order to paramterize the model.  This limits the application of these models to lakes with existing and, more importanly, available water quality data.  To expand our ability to predict in lakes without water quality data, we take advantage of the availability of a large national lakes water quality database, land use/land cover data, lake morphometry data, other universally available data, and modern data mining approaches to build and assess models of lake tropic state that may be more universally applied.  We use random forests and random forest variable selection to identify variables to be used for predicting trophic state and we compare the performance of two models of trophic state (as determined by chlorophyll *a* concentration).   The first model estimates trophic state with *in situ* as well as universally available data and the second model uses universally available data only. For each of these models we used three separate trophic state categories, for a total of six models. Overall accuracy for the *in situ* and universal data models ranged from xx% to xx% and xx, xx, and xx described the most variation in trophic state.  For the universal data only models, Overall accuraccy ranged from xx% to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it is believed that the presence and abundance of cyanobacteria is strongly associated with trophic state.  To test this we examine the association between estimates of cyanobacteria biovolume and the measured and predicted trophic state.  Expanding these preliminary results to include cyanobacteria taxa indicates that cyanobacteria are significantly more likely to be found in highly eutrophic lakes. These results suggest that predictive models of lake trophic state may be improved with additional information on the landscape surrounding lakes and that those models provide additional information on the presence of potentially harmful cyanobacteria taxa. 
\end{Abstract}
\end{singlespace}

#Introduction
Productivity in lentic systems is often categorized across a range of tropic states (e.g. the tropic continuum) from early succesional (i.e. oligotrophic)to late successional lakes (i.e. hypereutrophic) [@carlson1977trophic].  Lakes naturally occur across the range of trophic state and higher primary productivity is not necessarily a predictor of poor ecological condition.  Lakes that are naturally oligotrophic occur in nutrient poor areas or have a more recent geologic history.  These lakes are often found in higher elevations, have clear water, and are often favored for drinking water or direct contact recreation (e.g. swimming).   Lakes with higher productivity (e.g. eutrophic lakes) have greater nutrient loads, tend to be less clear, have greater density of aquatic plants, and often support more diverse and abundant fish communities.  Lakes will naturally shift to higher trophic states but this is a slow process.  Given this fact, monitoring trophic state allows the identification of rapid shifts in trophic state or locating lakes with unusually high productivity (e.g. hypereutrophic).  These cases are indicative of lakes under greater anthropogenic nutrient loads, also known as cultural eutrophication, and are more likely to be at risk of fish kills, fouling, and harmful algal blooms[@smith1998cultural;@smith1999eutrophication;@smith2006eutrophication].  Given the association between trophic state and many ecosystem services and disservices, being able to model trophic state could allow for estimating trophic state in unmonitored lakes and provide a first cut at identifying lakes with the potential for harmful algal blooms and other problems associated with cultural eutrophication.

@@

##Data and Study Area
We utilize four primary sources of data for this study,the National Lakes Assessment (NLA), the National Lake Cover Dataset (NLCD), modeled lake morphometery,  and estimated cyanobacteria biovolumes [@usepa2009national;@homer2004development;@xian2009updating;@hollister2010volume;@hollister_predicting_2011;@lakemorpho2014;@beaulieu2013nutrients].  All datasets are national in scale and provide a unique snapshot view of the condition of lakes in the United States'.

@@

##Predicting Trophic State with Random Forests
Random forest is a machine learning algorithm that aggregates numerous decision trees in order to obtain a consensus prediction of the response categories [@breiman2001random].  Bootstrapped sample data is recursively partitioned according to a given random subset of predictor variables and completely grown without pruning.   With each new tree, both the sample data and predictor variable subset is randomly selected.

@@

##Model Details
Using a combination of the `varSelRF` and `randomForest` we ran models for six combinations of variables and trophic state classifications.  These combinations included different combinations of the Chlorphyll *a* trophic states (Table \ref{tab:trophicStateTable}) along with all variables and the GIS only variables (i.e. no *in situ* infromation).  The six model combinations were:

@@

#Results
```{r figSetup,echo=FALSE,eval=TRUE,cache=TRUE}
#create list of all selected variables and assign a color
vars<-ls(pattern="_vars")
all_vars<-vector()
for(i in vars){
  all_vars<-c(all_vars,get(i))
}
all_vars<-unique(all_vars)

@@

##Model 1: 4 Trophic States ~ All Variables
```{r ALL_TS4, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts4<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_4,numCore=4,100,
    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="all_ts4",time=TRUE)
```
